Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-07-08 introduction.tex
Petite modif intro (2)

In this paper, we present a uniform framework for weighted parsing over infinite input alphabets.
\philippe{Tu fais une différence entre model et automata?}
It is based on \emph{symbolic weighted} finite states language models (\swM),
generalizing the Boolean guards of~\SA %in transitions
into functions into an arbitrary semiring,
and generalizing also~\WA, by handling infinite alphabets, see Figure~\ref{fig:hierarchy}.

@@

In short, a transition rule $q \xrightarrow{\phi} q'$ from state $q$ to $q'$ of a \swM,
%input symbols are variables, %(parameters)
is labeled by a function $\phi$ associating to every input symbol $a$ a weight value $\phi(a)$
in a semiring domain.
%and the weight associated to a transition rule is a function of these variables.
\philippe{This sentence (symbols as variables) is not immediately clear to me. Maybe a short example or intuition?}
\florent{modified}
%
\begin{figure}
\centering
\begin{tikzpicture}
\node (NFA) at (0,4) {%
  \(
  \begin{array}{c}
  \mathsf{FA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{B}\\
  q \xrightarrow{a} q' \quad  a \in \Sigma_\mathsf{fin}
  \end{array}
  \)
};
%
\node (WA) at (-2.5,2) {%
  \(
  \begin{array}{c}
  \mathsf{WA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{S}\\
  q \xrightarrow{a, w} q' \quad  a \in \Sigma_\mathsf{fin}, w \in \mathbb{S}
  \end{array}
  \)
};
%
\node (SA) at (2.5,2) {%
  \(
  \begin{array}{c}
  \mathsf{SA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{B}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{B}
  \end{array}
  \)
};
%
\node (SWA) at (0,0) {%
  \(
  \begin{array}{c}
  \mathsf{SWA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{S}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{S}
  \end{array}
  \)
};
\draw[->] (NFA)--(WA);
\draw[->] (NFA)--(SA);
\draw[->] (WA)--(SWA);
\draw[->] (SA)--(SWA);
%\begin{array}{c} \mathsf{NFA} : \Sigma^* \to \mathbb{B} \end{array}
\end{tikzpicture}
\caption{Classes of Symbolic/Weighted Automata.
$\Sigma_\mathsf{fin}$ is a finite alphabet,
$\Sigma_\mathsf{inf}$ is a countable alphabet,
$\mathbb{B}$ is the Boolean algebra,
$\mathbb{S}$ is a commutative semiring,
$q \xrightarrow{\dots} q'$ is a transition between states $q$ and $q'$.}
\label{fig:hierarchy}
\end{figure}
%
%This approach is close to the case of
%Symbolic Automata (SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM},
%except that the domain for weight values is not restricted to be Boolean,
%like for the guards in the rules of SA,
%but can be an arbitrary commutative semiring (assuming some restrictions).
%
\philippe{Tu veux dire: les modèles formels que tu combines?}
The models presented here are finite automata called symbolic-weighted (\SWA),
transducers (\SWT), and pushdown automata
with a visibly restriction~\cite{AlurMadhusudan09nested} (\SWVPA).
%\wrt an edit distance
The latter model of automata operates on \emph{nested words}~\cite{AlurMadhusudan09nested},
a structured form of words parenthesized with markup symbols,
corresponding to a linearization of trees.
In the context of parsing, they can represent (weighted) AST of CF grammars.
More precisely, a \SWVPA $A$ associates a weight value $A(t)$ % \in \Semiring$
to a given nested word $t$, which is the linearization of an AST. %representing a parse tree.
%
On the other hand,
a \SWT can define a distance $T(s, t)$ between finite words $s$ and $t$
over infinite alphabets. % following~\cite{Mohri03EDWA}.
Then, the \emph{SW-parsing} problem aims at %computing
finding $t$ minimizing
$T(s, t) \otimes A(t)$ (\wrt the ranking defined by $\oplus$), given an input word $s$.
The latter value is called the distance between $s$ and $A$ in~\cite{Mohri03EDWA}.
%

Like weighted-parsing
methods~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing},
our approach proceeds in two steps,
based on properties of the \swM.
The first step is an intersection
(Bar-Hillel construction~\cite{GruneJacobs08parsing})
\florent{chap. intersection in~\cite{GruneJacobs08parsing}}
where, given a \SWT $T$, a \SWVPA $A$, and an input word $s$,
a \SWVPA $A_{T, s}$ is built, such that for all $t$, $A_{T, s}(t) = T(s, t) \otimes A(t)$.
\philippe{The notation  $A_{T, s}$ has  not been introduced so far. It is not clear why $T$ is a parameter there}
In the second step, a best AST $t$ is found by applying to $A_{T, s}$
a best search algorithm similar to the shortest distance
in graphs~\cite{Mohri02semiring,Huang05kbest}.
%In expressiveness, they are equivalent to weighted CFG.
%and can be used in a general approach for parsing over infinite input alphabets.
%
%Let $A$ be a \SWVPA, associating $A(w) \in \Semiring$
%to a given a nested word $w$ (representing a parse tree),
%and let a \SWT compute a distance $d$, in $\Semiring$,
%between 2 strings over respectively an infinite input alphabet and the
%same (infinite) alphabet of $A$.
%Then, the problem of Symbolic Weighted Parsing is,
%given an input string $s$, to find a nested word $w$ minimizing
%(according to the ranking defined by $\oplus$)
%the distance $d(s, w) \otimes A(w)$ between $s$ and $A$,
%as defined in~\cite{Mohri03EDWA}.

>>>>

In this paper, we present a uniform framework for weighted parsing over infinite input alphabets.
It is based on \emph{symbolic weighted} finite states language models (\swM),
generalizing the Boolean guards of~\SA %in transitions
into functions into an arbitrary semiring,
and generalizing also~\WA, by handling infinite alphabets, see Figure~\ref{fig:hierarchy}.

@@

In short, a transition rule $q \xrightarrow{\phi} q'$ from state $q$ to $q'$ of a \swM,
%input symbols are variables, %(parameters)
is labeled by a function $\phi$ associating to every input symbol $a$ a weight value $\phi(a)$
in a semiring domain.
%and the weight associated to a transition rule is a function of these variables.
%
\begin{figure}
\centering
\begin{tikzpicture}
\node (NFA) at (0,4) {%
  \(
  \begin{array}{c}
  \mathsf{FA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{B}\\
  q \xrightarrow{a} q' \quad  a \in \Sigma_\mathsf{fin}
  \end{array}
  \)
};
%
\node (WA) at (-2.5,2) {%
  \(
  \begin{array}{c}
  \mathsf{WA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{S}\\
  q \xrightarrow{a, w} q' \quad  a \in \Sigma_\mathsf{fin}, w \in \mathbb{S}
  \end{array}
  \)
};
%
\node (SA) at (2.5,2) {%
  \(
  \begin{array}{c}
  \mathsf{SA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{B}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{B}
  \end{array}
  \)
};
%
\node (SWA) at (0,0) {%
  \(
  \begin{array}{c}
  \mathsf{SWA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{S}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{S}
  \end{array}
  \)
};
\draw[->] (NFA)--(WA);
\draw[->] (NFA)--(SA);
\draw[->] (WA)--(SWA);
\draw[->] (SA)--(SWA);
%\begin{array}{c} \mathsf{NFA} : \Sigma^* \to \mathbb{B} \end{array}
\end{tikzpicture}
\caption{Classes of Symbolic/Weighted Automata.
$\Sigma_\mathsf{fin}$ is a finite alphabet,
$\Sigma_\mathsf{inf}$ is a countable alphabet,
$\mathbb{B}$ is the Boolean algebra,
$\mathbb{S}$ is a commutative semiring,
$q \xrightarrow{\dots} q'$ is a transition between states $q$ and $q'$.}
\label{fig:hierarchy}
\end{figure}
%
%This approach is close to the case of
%Symbolic Automata (SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM},
%except that the domain for weight values is not restricted to be Boolean,
%like for the guards in the rules of SA,
%but can be an arbitrary commutative semiring (assuming some restrictions).
%
The framework relies on several language models:
finite automata called symbolic-weighted (\SWA),
transducers (\SWT), and pushdown automata
with a visibly restriction~\cite{AlurMadhusudan09nested} (\SWVPA).
%\wrt an edit distance
The latter model of automata operates on \emph{nested words}~\cite{AlurMadhusudan09nested},
a structured form of words parenthesized with markup symbols,
corresponding to a linearization of trees.
In the context of parsing, they can represent (weighted) AST of CF grammars.
More precisely, a \SWVPA $A$ associates a weight value $A(t)$ % \in \Semiring$
to a given nested word $t$, which is the linearization of an AST. %representing a parse tree.
%
On the other hand,
a \SWT can define a distance $T(s, t)$ between finite words $s$ and $t$
over infinite alphabets. % following~\cite{Mohri03EDWA}.
Then, the \emph{SW-parsing} problem aims at %computing
finding $t$ minimizing
$T(s, t) \otimes A(t)$ (\wrt the ranking defined by $\oplus$), given an input word $s$.
The latter value is called the distance between $s$ and $A$ in~\cite{Mohri03EDWA}.
%

Like weighted-parsing
methods~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing},
our approach proceeds in two steps,
based on properties of the \swM.
The first step is an intersection
(Bar-Hillel construction~\cite{GruneJacobs08parsing})
\florent{chap. intersection in~\cite{GruneJacobs08parsing}}
where, given a \SWT $T$, a \SWVPA $A$, and an input word $s$,
a \SWVPA $A_{T, s}$ is built, such that for all $t$, $A_{T, s}(t) = T(s, t) \otimes A(t)$.
In the second step, a best AST $t$ is found by applying to $A_{T, s}$
a best search algorithm similar to the shortest distance
in graphs~\cite{Mohri02semiring,Huang05kbest}.
%In expressiveness, they are equivalent to weighted CFG.
%and can be used in a general approach for parsing over infinite input alphabets.
%
%Let $A$ be a \SWVPA, associating $A(w) \in \Semiring$
%to a given a nested word $w$ (representing a parse tree),
%and let a \SWT compute a distance $d$, in $\Semiring$,
%between 2 strings over respectively an infinite input alphabet and the
%same (infinite) alphabet of $A$.
%Then, the problem of Symbolic Weighted Parsing is,
%given an input string $s$, to find a nested word $w$ minimizing
%(according to the ranking defined by $\oplus$)
%the distance $d(s, w) \otimes A(w)$ between $s$ and $A$,
%as defined in~\cite{Mohri03EDWA}.
